---
title: "Lintr And Covr"
author: Jim Hester
date: August 26th, 2015
output:
  ioslides_presentation:
    widescreen: true
vignette: >
  %\VignetteIndexEntry{Covr Exercises}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Who Is Jim Hester? ##

  - [Bioconductor](https://www.bioconductor.org) Core Member - Remote
  - Previous
    - [Explorys](www.explorys.com) - Data Scientist - Cleveland
    - [Cleveland Clinic](www.ccf.org) - Bioinformatician - Cleveland
  - R User - 2008
  - Contributor - [devtools][], [ggplot2][], [dplyr][], [rmarkdown][], [plyr][], [knitr][],
    [tidyr][], [roxygen][], [httr][], [ggvis][], [markdown][]
  - Package Author - [covr][], [lintr][], [rex][], [gmailr][], [knitrBootstrap][], ...

[covr]: https://github.com/jimhester/covr
[dplyr]: https://github.com/hadley/dplyr/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[ggplot2]: https://github.com/hadley/ggplot2/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[ggvis]: https://github.com/rstudio/ggvis/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[gmailr]: https://github.com/jimhester/gmailr
[httr]: https://github.com/hadley/httr/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[knitr]: https://github.com/yihui/knitr/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[lintr]: https://github.com/jimhester/lintr
[primerTree]: https://github.com/jimhester/primerTree
[rex]: https://github.com/jimhester/rex
[tidyr]: https://github.com/hadley/tidyr/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[rmarkdown]: https://github.com/rstudio/rmarkdown/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[plyr]: https://github.com/hadley/plyr/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[markdown]: https://github.com/rstudio/markdown
[devtools]: https://github.com/hadley/devtools/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[roxygen]: https://github.com/klutometis/roxygen/pulls?q=is%3Apr+author%3Ajimhester+is%3Aclosed
[knitrBootstrap]: https://github.com/jimhester/knitrBootstrap

----

- [lintr][] - Static Code Analysis / Linting
- [covr][] - Code Coverage Analysis / Reporting

## Lintr Motivation

  - Other Languages
    - [flake8](https://pypi.python.org/pypi/pyflakes), [pylint](http://en.wikipedia.org/wiki/Pylint), [PerlTidy](http://en.wikipedia.org/wiki/PerlTidy), [jshint](http://jshint.org/docs/), [clang](http://clang.llvm.org/), [lint](http://www.unix.com/man-page/FreeBSD/1/lint)

  - On The Fly
    - Vim - [Syntastic](https://github.com/scrooloose/syntastic)
    - Emacs - [flycheck](https://github.com/flycheck/flycheck)
    - Sublime Text - [SublimeLinter](http://www.sublimelinter.com/)

  - Style / Syntax
    - Own/others contributions (pull requests)
    - Hadley style (<http://r-pkgs.had.co.nz/style.html>)

  - Synergy
    - Use case for [rex](https://github.com/kevinushey/rex)


## Existing Alternatives

  - [lint](https://github.com/halpo/lint) - *Andrew Redd*
    - Stagnated development, Feb 5, 2013
    - Slow performance?
    - Limited set of linters

  - [svTools](http://cran.r-project.org/web/packages/svTools/index.html) - *Philippe Grosjean, Romain Francois*
    - Major development 2010, last bugfix release Mar 2014
    - many features un-related to linting
    - linting implementation wraps codetools functions

  - [codetools](http://cran.r-project.org/web/packages/codetools/index.html) - *Luke Tierney*
    - Used internally by R CMD check
    - Checks for possible usage errors, not style

  - [shinyapps](https://github.com/rstudio/shinyapps/issues/103) - *Kevin Ushey*
    - Shiny app website only


## Demo #
  - R terminal
  - Vim
  - Emacs
  - Sublime Text
  - RStudio
  - Travis-CI

## Configuration #
  - Some linters have arguments
    - `line_length_linter(120)`
  - Arguments to lint function
  - Per project configuration file
  - Exclusions
    - Whole files
    - Line
    - Ranges
    - In-source and outside
  - camelCase vs snake_case vs ambiguous.case

## Implementation #
  - `base::parse()`
    - Parse errors -> errors
  - `utils::getParseData()`

```{r}
sf <- srcfile("bad.R")
p <- try(parse(text=readLines(sf$filename),
               srcfile=sf,
               keep.source = TRUE))
(pd <- getParseData(sf))
```

## Example linter #
```r
assignment_linter <- function(source_file) {
  lapply(ids_with_token(source_file, "EQ_ASSIGN"),
    function(id) {
      parsed <- source_file$parsed_content[id, ]
      Lint(
        filename = source_file$filename,
        line_number = parsed$line1,
        column_number = parsed$col1,
        type = "style",
        message = "Use <-, not =, for assignment.",
        line = source_file$lines[parsed$line1]
        )
    })
}
```

## Example linter ##
```r
trailing_whitespace_linter <- function(source_file) {
  res <- re_matches(source_file$lines,
    rex(capture(name = "space", some_of(" ", regex("\\t"))), or(newline, end)),
    global = TRUE,
    locations = TRUE)

  lapply(seq_along(source_file$lines), function(itr) {

      mapply(
        FUN = function(start, end) {
          if (is.na(start)) {
            return()
          }
          line_number <- names(source_file$lines)[itr]
          Lint(
            filename = source_file$filename,
            line_number = line_number,
            column_number = start,
            type = "style",
            message = "Trailing whitespace is superfluous.",
            line = source_file$lines[as.character(line_number)],
            ranges = list(c(start, end)),
            linter = "trailing_whitespace_linter"
            )
        },
        start = res[[itr]]$space.start,
        end = res[[itr]]$space.end,
        SIMPLIFY = FALSE
        )
  })
}
```

## Implementation #

  - linting speed an issue
    - ~ 20 seconds to lint lintr, 60 files
  - Caching
    - Experimental
    - Per expression
    - Cache dependencies still a work in progress

## Future Directions #

  - Bioconductor linters
    - Very close to hadley style
    - camelCase
    - no spaces in argument lists (a=b)
  - Google Style
  - Improve Performance
    - C/C++ helper utility functions?
  - Automatic reformatting/tidying ([formatR](http://cran.r-project.org/web/packages/formatR/index.html))
    - False positives

## Lintr #

  - Integrated with common editors
  - Works with both packages and scripts
  - Style, syntax and potential usage errors
  - Easy to understand output
  - Configurable
  - Lintr questions?

# Covr #
## What is it? #

  - Test Coverage - How much of my code is run by tests?
  - Mid-December 2014
  - Test/Example/Vignette Coverage
    - R Code
    - Compiled C/C++/Fortran Code

## Motivation ##

  - What tests are needed?
  - Gauge of package robustness - [Shields](http://shields.io/) / Badges
  - Other languages coverage tools
    - [gcov](https://gcc.gnu.org/onlinedocs/gcc/Gcov.html)
    - [JCov](https://wiki.openjdk.java.net/display/CodeTools/jcov)
    - [coverage](https://pypi.python.org/pypi/coverage)
    - [simplecov](https://rubygems.org/gems/simplecov)
  - Services
    - [coveralls.io](https://coveralls.io/)
    - [codecov.io](https://codecov.io/)

## Existing Alternatives ##

  - [R-coverage](http://r2d2.quartzbio.com/posts/r-coverage-docker.html) by Karl Forner
    - Modify R source and add instrumentation
    - Requires patching and recompiling R source
  - [testCoverage](https://github.com/MangoTheCat/testCoverage) by Tom Taverner, Chris Campbell, Suchen Jin
    - Alternate parser
    - Complicated implementation
    - No S4 support
    - Limited output formats
    - Challenging usage instructions

## Demo ##

  - R terminal
  - Shiny application
  - Coveralls.io
  - Codecov.io

## Configuration ##

  - Exclusions (Cheating!)
    - Whole files
    - By Line
    - Ranges
    - In-source annotations and `exclusion` argument.

# Implementation #
## R's Abstract Syntax Tree ##

  - [How it works](https://github.com/jimhester/covr/blob/master/vignettes/how_it_works.Rmd) vignette
  - Walk the Abstract Syntax Tree
    - If a call with srcref
      - Add a trace function before call
      - Perform the call

----

```{r eval = FALSE}
fun <- function(x, ...) {
  recurse <- function(y) {
    lapply(y, fun, f = f, ...)
  }

  if (is.atomic(x) || is.name(x)) {
  } else if (is.call(x)) {
    as.call(recurse(x))
  } else if (is.function(x)) {
    formals(x) <- fun(formals(x), ...)
    body(x) <- fun(body(x), ...)
    x
  } else if (is.pairlist(x)) {
    as.pairlist(recurse(x))
  } else if (is.expression(x)) {
    as.expression(recurse(x))
  } else if (is.list(x)) {
    recurse(x)
  } else {
    stop("Unknown language class: ", paste(class(x), collapse = "/"),
      call. = FALSE)
  }
}
```

## Modify Calls ##

  - How to insert function without changing output?
  - Braces evaluate expressions, return result of last

```{r}
identical({ 1 + 2; 3 + 4 }, `{`(1 + 2, 3 + 4))
```

```{r, eval = FALSE}
`{`(count(), as.call(recurse(x)))
```

## Source References ##

  - Where in source a call is from?
  - srcref
    - `option(keep.source = TRUE)`
    - srcref attribute attached to each call

----

```{r}
f1 <- function(x) {
  x <- x + 1
  y <- x
  y
}

covr:::trace_calls(f1)
```

## Replacing functions ##

  - Replace all references with modified versions.
  - `testthat::with_mock()`
    - C function replaces function pointer
    - stores original definition (reversible)

## S4 Methods ##

  - Normal functions defined directly in package namespace
  - S4 methods are defined in an environment based on their generic

```{r eval = FALSE}
replacements_S4 <- function(env) {
  generics <- getGenerics(env)

  unlist(recursive = FALSE,
    Map(generics@.Data, generics@package, USE.NAMES = FALSE,
      f = function(name, package) {
      what <- methodsPackageMetaName("T", paste(name, package, sep = ":"))

      table <- get(what, envir = env)

      lapply(ls(table, all.names = TRUE), replacement, env = table)
    })
  )
}
```

## Compiled Code ##

  - Gcov
    - built into gcc and clang
      - `-fprofile-arcs -ftest-coverage`
      - `-O0`
    - Need to override default and package Makevars
      - `PKG_CFLAGS` puts optimization _before_ default `-O2`
    - Temporarily point to different global Makevars. (retain `~/R/Makevars` values)
    - No results until process terminated
      - Call R subprocess

## Running Tests ##

  - `base::source2` on `tests/*.[Rr]`
  - Test framework agnostic

## Running Vignettes ##

  - `knitr::knit(tangle = TRUE)` # generate R script
  - Run with `base::source2`

## Running Examples ##

  - tools:::.createExdotR # generate R script from `.Rd`
  - script massaged to use temp directory and not quit.

## Coverage Services ##

  - Track coverage over time
  - Report on coverage drops from contributions.
    - Coveralls.io
    - Codecov.io

  - Expect a JSON file
    - coverage per line
    - source code per line
  - Travis CI
    - Travis Job ID

## Covr Usage / Shields ##

  - Designed for [README.md](https://github.com/jimhester/covr/blob/master/README.md) (GitHub)
  - Simple indicator of test breadth
  - Currently used in nearly [200 projects on GitHub](https://github.com/search?l=yaml&p=1&q=in%3Afile+.travis.yml+jimhester%2Fcovr&ref=simplesearch&type=Code&utf8=%E2%9C%93)
    - https://github.com/Rdatatable/data.table
    - https://github.com/yihui/knitr
    - https://github.com/hadley/testthat
  - Since Mid-June Bioconductor Packages ~ 328 with unit tests
    - https://codecov.io/github/Bioconductor-mirror
    - Package landing pages e.g (<http://bioconductor.org/packages/SummarizedExperiment/>)

## Future ##

  - Ideas?
  - Comments / Questions?
