---
output:
  revealjs::revealjs_presentation:
    reveal_plugins: ["zoom"]
    self_contained: false
    theme: "solarized"
    center: true
    smart: false
    slide_level: 2
---

# R Syntax Extensions

Jim Hester - RStudio

---

- Most proposals have working implementations linked
- Examples from other languages
- Proposed R Solution

# [Raw strings](https://github.com/wch/r-source/compare/trunk...jimhester:feature-raw_string)

## Example - Python
```{python}
print(r"a / raw string with \ some 'quotations'")
print(r'Python has four forms to handle "different" quotes')
print(r"""Including the 'triple' "quoted" form""")
```

## Example - Perl
```{perl}
print q{Perl \ lets you delimit 'with' "paired" delimiters };
print q#and alternatively any 7-bit \Ascii character#
```

## Proposed Syntax
```{r}
# Raw strings are very useful in regular expressions
gsub(r"\\", "/", r"\a\new\string")

# Or windows file paths
r'C:\foo\bar'
```

# [String interpolation](https://github.com/tidyverse/glue)

## Example - Bash

```{bash}
value="foo"
echo "value = $value"
```

```{r}
foo <- 1
bar <- 2
paste0("foo = '", foo, "'
bar = '", bar, "'")

sprintf("foo = '%s'\nbar = '%s'", foo, bar)
```

## Proposed Syntax

```{r}
#install.package("glue")
g <- glue::glue

g("foo = '{foo}'
   bar = '{bar}'")

g("1 + 1 = {a <- 1; a + 1}")

g("foo = '{a}'
   bar = '{b}'",
  a = format(foo, nsmall = 2),
  b = bar + 2)

gd <- glue::glue_data
mtcars$name <- rownames(mtcars)
gd(head(mtcars), "The {name} has {mpg} mpg and {hp} hp.")
```

## Vs `sprintf()`
  - No explicit typing
  - Implicitly documented
  - Maintain variable locality

# [Unary Specials](https://github.com/wch/r-source/compare/trunk...jimhester:feature-unary_operators)

## Proposed Syntax

```{r}
`%chr%` <- as.character

%chr% 1
```

## Advantages

- Allows user defined binary `%foo%` in unary contexts.
- Most useful for operator that make sense in binary and unary contexts.

---

```{r}
`%g%` <- function(x, y) {
  env <- parent.frame()
  if (missing(y)) {
    eval(bquote(glue::glue(.(x), .envir = env)))
  } else {
    eval(bquote(glue::glue_data(x, .(y), .envir = env)))
  }
}
```

---

```{r}
x <- 10
%g% "x = {x}"

head(mtcars) %g% "The {name} has {mpg} and {hp} hp."
```

---

```{r}
library("magrittr")
mtcars %>% head %>% summary
sum <- . %>% head %>% summary
sum
```

---

```{r}
# devtools::install_github("jimhester/magrittr@unary")
sum <- %>% head %>% summary
sum
```

# [Brackets for List Instantiation](https://github.com/wch/r-source/compare/trunk...jimhester:feature-brackets)

## Example - Ruby

```{ruby}
days = [
  "Monday",
  "Tuesday",
  "Wednesday",
  [1, 2, 3],
]
print(days)
```

## Proposed Syntax
```{r}
days <- [
  "Monday",
  "Tuesday",
  "Wednesday",
  [1, 2, 3]
]
days
```

# `foreach` and Generator functions

## Example - Python

```{python}
import warnings
def yrange(n):
    i = 0
    while i < n:
        yield i
        i += 1

y = yrange(3)
print(y)
print(y.next())
print(y.next())
print(y.next())
try:
    print(y.next())
except StopIteration as e:
    print("Iterator done")

# Within a loop
for i in yrange(3):
    print(i)
```

---

```{r}
yrange <- function(n) {
  i <- 1
  function() {
    if (i <= n) {
      i <<- i + 1
      return(i - 1)
    }
    return(FALSE)
  }
}
y = yrange(3)
y()
y()
y()
y()
```

---

```{r}
y <- yrange(3)
while (i <- y()) {
  print(i)
}
```

## Proposed Syntax

```r
foreach (i in yrange(3)) {
  print(i)
}

#> [1] 1
#> [1] 2
#> [1] 3
```

- Implement with Condition objects
- Symmetry with current `for` syntax.

# [Type Annotations](https://cran.r-project.org/package=types)

## Example - Python 3

```{python3}
def greeting(name: str) -> str:
    hi: str = 'Hello '
    return hi + name

print(greeting("Jim"))
```

## Proposed Syntax
```{r, error = TRUE}
# install.packages("types")
suppressPackageStartupMessages(library(types))
greeting <- function(name = ? str) {
  hi <- "Hello " ? str
  paste0(hi, name)
} ? str
greeting()
greeting("Jim")
```

- [mypy](http://mypy-lang.org/) Static type checker for Python.
- Backwards compatible - no parser modifications
- [typeCheck](https://github.com/jimhester/typeCheck) - Add runtime checks based on type annotations.
- Future
  - Auto generate documentation
  - Static type checker
  - Compiler optimizations

# [Native pipe operator](https://github.com/wch/r-source/compare/trunk...jimhester:feature-native_pipe)

## Example - sh

```{sh}
ps aux | grep -w R | awk '{print $2}' | head
```

```{r, error = TRUE}
library(magrittr)
fail <- function(...) stop("fail")
mtcars %>% lapply(fail) %>% unlist()
traceback()
#> 14: stop("fail") at #1
#> 13: FUN(X[[i]], ...)
#> 12: lapply(., fail)
#> 11: function_list[[1L]](value)
#> 10: unlist(.)
#> 9: function_list[[1L]](value)
#> 8: freduce(value, `_function_list`)
#> 7: Recall(function_list[[1L]](value), function_list[-1L])
#> 6: freduce(value, `_function_list`)
#> 5: `_fseq`(`_lhs`)
#> 4: eval(quote(`_fseq`(`_lhs`)), env, env)
#> 3: eval(quote(`_fseq`(`_lhs`)), env, env)
#> 2: withVisible(eval(quote(`_fseq`(`_lhs`)), env, env))
#> 1: mtcars %>% lapply(fail) %>% unlist()
```

## Proposed Syntax

```{r, error = TRUE}
mtcars >> head
```

```{r}
mtcars >> head()
```

---

```{r}
mtcars >> 
  subset(cyl == 4) >>
  lm(mpg ~ disp, data = _)

quote(mtcars >> 
  subset(cyl == 4) >>
  lm(mpg ~ disp, data = _))
```

```{r, error = TRUE}
mtcars >> lapply(fail) >> unlist()
traceback()
#> 4: stop("fail") at #1
#> 3: FUN(X[[i]], ...)
#> 2: lapply(mtcars, fail)
#> 1: unlist(mtcars |> lapply(fail))
```

---

- Better error messages.
- Much better tracebacks on failure.
- Consistent use of `_` as the pipe placeholder instead of `.`.

# [Trailing Commas](https://github.com/wch/r-source/compare/trunk...jimhester:feature-trailing_arguments)

## Example - C99
- C (99),C++, perl, python, ruby, julia, ...
```c
enum {
  foo,
  bar,
  baz,
} e;
```

---

```{r}
c(
  1,
  3,
  2,
)
#> Error in c(1, 2, 3, ) : argument 4 is empty
c(1,
  2,
  3,
  NULL)
```

---

```{r}
list(1,
     2,
     3,
     NULL)
```
```{r}
c(1,
  2,
  3,
  list())
```

## Proposed Syntax
```{r}
c(
  1,
  2,
  3,
)
```

# [Compact anonymous functions](https://github.com/jimhester/lambda#readme)

## Example - C++11

```c
/* engine.opts = "-x c++ -std=c++11 -o lambda" */
#include <iostream>
#include <functional>

int main()
{
  auto f = [](int i) { return i + 4; };
  int num;
  std::cin >> num;
  std::cout << f(num) << '\n';
}
```

```{bash}
echo 0 | ./lambda
echo 10 | ./lambda
```

## Proposed Syntax

```{r}
# devtools::install_github("jimhester/lambda")
library("lambda")
fun <- f({.(x) + .(y = 1)})
fun(1)
fun(1, 2)

lapply(1:3, f({.(x) + .(y)}), y = 3)

x <- list(1, NULL, 2)
compact <- f(Filter(Negate(is.null), .(x)))
base::print.function(compact)
compact(x)
```

---

```{r}
`%f%` <- lambda::f
%f% { .(x) + .(y = 2) }
```

# [Triple-Quoted String Literals](https://cran.r-project.org/web/packages/glue/README.html)

```julia
str = """
         Hello,
           world.
         """
#> "Hello,\n  world.\n"
```

---

```{r}
# install.packages("glue")
t <- glue:::trim

t("hello")
t("
hello")
t("

hello")

t("
  Hello,
    World.
  ")
```

## Proposed syntax
```r
f <- function() {
  """
  Hello,
    World.
  """
}
```
# [Variable unquoting / tidy evaluation](http://dplyr.tidyverse.org/articles/programming.html)

## Example - SQL

```{r}
library("DBI")
con <- DBI::dbConnect(odbc::odbc(), "MySQL")
mtcars$name <- row.names(mtcars)
dbWriteTable(con, "mtcars", mtcars, overwrite = TRUE)
dbExecute(con, "SET @cyl = 8, @hp = 225")
dbGetQuery(con, "SELECT name FROM mtcars WHERE cyl = @cyl AND hp > @hp")
```

---

```{r}
suppressPackageStartupMessages(library(rlang))
var <- "x"
value = "y"
data.frame(exprs(!!var := !!value))
```

## Proposed Syntax
```r
var <- "x"
value = "y"
data.frame(@var = @value)
```

# Variable unpacking

## Example - Ruby

```{ruby}
data = ['a', 'b', 'c']
foo, bar, baz = data
puts(foo, bar, baz)
```

---

```{r}
`%<-%` <- function(lhs, rhs) {
  lhs <- substitute(lhs)[-1L]
  for (i in seq_along(lhs)) {
    assign(as.character(lhs[[i]]), rhs[[i]], envir = parent.frame())
  }
}
data <- c("a", "b", "c")
c(foo, bar, baz) %<-% data
foo; bar; baz
```

## Proposed syntax
```r
foo, bar, baz <- data
```
